Forage site distributions quantify collective foraging in honey bee colonies: Methods

### 2022-04-08 analysis/manuscript/methods.Rmd
updated methods

We make the initial assumption that there are $n$ different resources distributed within an environment according to a spatial Poisson point process with rate $\lambda_{I}$ for resource $I$. As a scout travels a small distance $dx$ it will cover an area of $c_sdx$, with $c_s$ depending upon the width of the area scanned, the detection rate and degree to which a scout's path covers new ground. The probability density of the distance travelled after which the first resource is discovered $x$ follows an exponential distribution with a rate derived from the intensity of the flowers within the environment: $\lambda_s e^{-\lambda_s x}$, where $\lambda_s = c_s \sum_{I = 1}^{n} \lambda_I$.

The number of repeats of a dance on the dance floor is dictated by profitability and is defined as the net energy efficiency per amount of energy spent: $\frac{Gain-Cost}{Cost}$. The gain is determined by the sucrose concentration of nectar from a flower, $q$, whilst the cost is determined by the distance, $x$. The energy gained, $G$, is therefore proportional to the quality $q$, whilst the cost increases linearly with distance. We can therefore refactor energy efficiency as $\frac{q}{1+\alpha x}-1$, where $\alpha$ represents the energy spend per distance relative to a fixed cost of travel. As bees should not be expected to dance for a negative profitability, we can derive a function to translate quality and distance into a measure of the number of dances on a dance floor: $\sigma(q,x) = [\frac{q}{1+\alpha x}-1]_+$, where the notation of $[x]_+ = max(x, 0)$. This function has the property that for a fixed $x$ there is a minimum quality $q = 1+\alpha x$ below which bees don't dance. The equivalent quality of two resources at distance $x1$ and $x2$ is $\frac{1+\alpha x2}{1+\alpha x1} = \frac{q2}{q1}$. The critical distance $\epsilon_{ij} = [\frac{1}{\alpha} \frac{qj - qi}{qi}+\frac{qj}{qi} x]_+$.

For a resource of type $I$ located at a distance $x$, the expected number of dances would be $c_s \lambda_I \sigma(qi, x) e^{-\lambda_s x}$. It follows that the expected total number of scout waggle dances on the dance floor is given by: $M_s = \sum_{I=1}^{n} \int_{0}^{\inf} \frac{c_s \lambda_I}{\lambda_s} \sigma(qi, x) \lambda_s e^{-\lambda_s x} dx$. The distribution of scout dances is: $\frac{\lambda_s e^{-\lambda_s x} \sum_{I=1}^{n} \frac{c_s \lambda_I}{\lambda_s} \sigma(qi, x)}{M_s}$.

Using the same logic we derive what the distribution of recruits will look like. If two locations differ in resource quality and distance then a hive can prefer a further location if it is sufficiently better and a lesser quality resource if it is sufficiently nearer. For a site with quality $q_I$ and at distance $x$ from the hive, for a second location with quality $q_j$ this threshold distance $\epsilon_{ij}$ satisfies $\sigma(q_i, x) = \sigma(qj, \epsilon_{ij}$. Consequently, the critical distance is an implicit function of distance and it follows from implicit differentiation that $\frac{d\epsilon_{ij}}{dq_j} = - \frac{d\sigma{q_j,x}/dq{j}}{d\sigma{q,x}/dx} > 0$ and $\frac{d\epsilon_{ij}}{dq_I} = \frac{d\sigma{q_i,x}/dq{i}}{d\sigma{q,x}/dx} < 0$. If $qi = qj$ then $\epsilon_{ij}(x) = x$ and it follows that if $q_I > q_j$ then $\epsilon_{ij} < x$ and if $q_I < q_j$ then $\epsilon_{ij} > x$.  As recruits get their information from scouts (as well as recruits) how well the scouts cover depends upon the number of scouts and how well their collective paths cover an area and the chance of discovery of each resource. To capture this we let the fraction of resources that can be discovered by a collection of scouts be $c_r$ and assume $0 \leq c_r \leq 1$. The probability that there is a not a food source discovered worth reporting on within a radius of $x$ $e^{-c_r \lambda_r \pi \epsilon_{ij}(x)^2}$, where $\lambda_r = \sum_I={1}^{n} \lambda_i$. This function also described the probability of having the nearest food source that the recruits will come out for discovered at least a distance of $x$ away.

@@

All models are fit using Maximum likelihood estimation [@Burnham2002] by summation of the log of the model function outlined in the methods section: model. The numerical optimisation routine is written in c++ and uses the Nelder-Mead simplex algorithm [@Nelder1965] implemented in the 'NLopt' library [@Johnson2020] and interfaced to R [@RCore2020] using 'Rcpp' [@Eddelbuettel2011].

Model parsimony is assessed using Akaike information criterion (AIC) [@Aho2014] and the model with the lowest AIC score is deemed to be the most parsimonious.

Goodness of fit is assessed using the Kolmorgorov-Smirnov (KS) test [@Goldstein2004] and implemented in R using the ks.boot function of the package Matching in R [@Sekhon2011].

@@

A key consideration is that our response variable is a proportion and thus lies on the interval $[0, 1]$. This makes it unsuitable for a standard linear regression. One option would be to use a generalized linear regression with a logit link and quasibinomial family. This method is most suited for when the response variable is an integer proportion, where the response variable is a division of two integers, such as the proportion of males within a population. Our measure is in principle an integer proportion as we measure the proportion of scouts, however, this value is an estimate derived from our model and so is continuous on the interval $[0, 1]$. An alternative method would be to transform the proportions on to a continuos scale using the $arcsin \sqrt{x}$ transformation. The most appropriate method would be to use a beta regression which is designed specifically for proportional response variables.

We used the R package plsRbeta [@Bertrand2013] to conduct the partial least squares analysis and performed a beta regression on the results using the R package betareg [@Cribari-Neto2010].

>>>>

To describe the distribution of dance durations on the dance floor we formulated a generic model the duration of waggle dances. In the model resource patches are randomly placed in the environment. Foragers scout for these patches. The rationale of the model is illustrated in Fig. 3: upon visiting a resource patch, foragers translate the profitability of a resource into the number of repeats of the dance. The number of repeats of the dance is a function of quality an distance. Recruits sample random dances and report the location of successful visits to resource patches on the dance floor. Through the feedback and over-representation of profitable resources on the dance floor recruits will converge to visiting the most profitable resource in vicinity of the hive. The distribution of dance durations is the superposition of scouting and recruiting trips.

As the resources patches are randomly placed in the environment, the distance after which the first resource is discovered approximately follows an exponential distribution (given by $\lambda e^{- \lambda x}$). Through the feedback mechanism that the dance floor provides, the colony can, collectively, locate the most profitable resource in its environment. For randomly placed resources in a two dimensional environment the distance to the nearest point is distributed according to a Rayleigh distribution (given by $2 \lambda \pi x e^{- \pi \lambda x^2}$) (Pyke 1978). Our simulation model shows that this describes the distances at which recruits visit resources well.  Knowing the distance distributions of scout and recruit trips we then assume that the a proportion $p$ of all trips are scout trips. With this information we can specify the distributions of distances on the dance floor (see Supplementary Material for details).

We implemented this in a full model which describes the distance distribution of the environment has $n$ different resource types (See Supplementary Material). However, in the full model the number of parameters increases with $2n$.  Even if the number of resources is low, it turned out to be cumbersome to estimate the parameters and the model tends to overfit. To facilitate estimation of the parameter $p$ we therefore used a simplified model to estimate the fraction of scout trips.

For the simplified the model we assumed that the the by assuming that the number of dances depends weakly on distance and there is a sizable quality differences between resources of a decent size and that there is a sizable intensity of the high quality resource. Foragers on scouting trips are more likely to report larger distances than foragers on recruiting trips. By linearising the function that translates the profitability into the number of waggle dance run for the largest distance and normalising, we arrive at simplified distribution for distances for scouting trips:

$$ f_s(x)=M_s^{-1} a_s b_s e^{-b_s a_s x}  \left[ 1-a_s x\right]_+ ,$$

where we used the shorthand ${x}_+=\max(x,0).$ The parameter $a_s^{-1}$ is the maximum reported distance by scouts, $a_s b_s$ is the intensity of resources found by scouts and $M_s=1-b_s^{-1} \left(1-e^{-b_s}\right)$ is factor that normalises the distribution.

Recruit trips will be predominantly to high quality resource. Only if the nearest high quality patch is very far away will there be a more profitable patch of lesser quality available, and this happens only rarely if the intensity of the best quality resource is sizable. After linearising the function that translates the profitability into the number of waggle dance run for short distances and normalising the distribution of distances reported from recruit trips in the simplified model is:

$$f_r(x)=M_r^{-1} 2 \pi a_r^2   b_r x e^{- \pi  b_r  (a_r x)^2}\left[1- a_r x\right]_+$$

where $M_r={1-\frac{erf\left(  \sqrt{\pi b_r}\right)}{2 \sqrt{b_r }}}$ is the normalisation factor, the parameter $a_r$ is the rate with which dances repeats depends on distance for recruit trips and $a_r^2 b_r$ is the intensity of high quality resources reported by recruited foragers.

The simplified distribution function is

$$P(\underline x= x )=p f_s(x)+ (1-p) f_r(x), $$ which we used for parameter estimation and model fittings.

@@

All models are fit using Maximum likelihood estimation [@Burnham2002] by summation of the log of the simplified distribution function outlined in the methods section: model. The numerical optimisation routine is written in c++ and uses the Nelder-Mead simplex algorithm [@Nelder1965] implemented in the 'NLopt' library [@Johnson2020] and interfaced to R [@RCore2020] using 'Rcpp' [@Eddelbuettel2011].

Model parsimony is assessed using Akaike information criterion (AIC) [@Aho2014] and Akaike weights. The model with the lowest AIC score is deemed to be the most parsimonious.

Goodness of fit is assessed using the Kolmorgorov-Smirnov (KS) test [@Goldstein2004] and implemented in R using the ks.boot function of the package 'Matching' in R [@Sekhon2011].

@@

As our estimated proportion of scouts is continuos on the interval $[0, 1]$ we performed a beta regression using the R package plsRbeta [@Bertrand2013] to conduct the partial least squares analysis and performed a beta regression on the results using the R package betareg [@Cribari-Neto2010]. As the betareg package only works on the open interval $(0, 1)$ the data, $x$, was transformed using the following equation: $\frac{x(n-1)+0.5}{n}, as outlined in the betareg package documentation. After analysis the data was back transformed to the original values for the plots in Fig 5.

For the jackknifed resampling we iterated through the each site and removed it from the pool of data and then ran the PLS as described above, recoding the loadings for each iteration (see Supplementary Material for loadings with each site removed). The PLS loadings for each land-use type are plotted as a box plot in Fig 5. to show the spread of these variable types.
